Quantitative Benchmarking for Automatically Generated Conjectures

### 2018-02-26 TEBenchmarkPaper/article.tex
Use et al consistently

@@

A major challenge for both ATF and MTE is choosing how to narrow down
the set of generated conjectures to those deemed ``interesting'', since this is
an imprecise term with many different interpretations. For example, all existing
approaches agree that simple tautologies are ``uninteresting'', but differ when
it comes to more complex statements.  Colton {\em et al.} give a survey of
systems for theory formation and exploration, and their notions of
``interestingness'' for concepts and conjectures, identifying six notions in
total~\cite{colton2000notion}. These notions, as applied to conjectures, are:

@@

{\bf Applicability} depends on the number of models in which a
conjecture holds. The system of Bagai et al conjectures the
non-existence of objects, and hence favours statements with zero
applicability. Other systems treat applicability as a positive
aspect: the more applicable the statement, the more interesting it is.

@@

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                     &
    \multirow{2}{*}{\textbf{Conjecture Types}}            &
    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{            &                    & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{GT          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti    & \tINE              &   X &   X &   X &     &   X &   X}
    \tRow{Bagai et al & \tNE               &     &   X &     &   X &   X &    }
    \tRow{HR          & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec   & \tEQ               &   X &   X &     &   X &   X &    }
    \tRow{Speculate   & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy     & \tEQ               &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme   & \tEQ               &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE systems from \cite{colton2000notion}, extended
    to those compared in \cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec). The interestingness measures
    are \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

We summarise the criteria in Table 1 and state which were used in key
ATF/MTE systems. The entries for AM, GT, Graffiti, Bagai et al. and HR are based
on this survey (the latter is Colton's own system). We extend this analysis to
some more recent systems in the remaining rows (based on our understanding of
their function): QuickSpec, Speculate~\cite{braquehais2017speculate}, IsaCoSy
and IsaScheme.

@@

%IsaScheme
Monta{\~n}o-Rivas {\em et al.} have implemented a scheme-based
approach to MTE in their IsaScheme system
\cite{MontanoRivas2011}. Schemes are higher-order formulae which can
be used to generate new concepts and conjectures; variables within the
scheme are instantiated automatically and this drives the invention
process.  For instance, in the theory of natural numbers, given the
concepts of successor ($suc$), addition ($+$) and zero ($0$),
IsaScheme can use the following scheme to invent the concept of
multiplication:

@@

Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical development
\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly proposed
metric. Methods for finding out what that means have come from neuro-scientists,
such as the study by Zeki et al in which they test whether mathematicians'
experiences of abstract beauty correlates with the same brain activity as
experiences of sensory beauty \cite{zeki}; computer scientists, such as Colton's
metrics above which are proposed based largely on a computer scientist's
``intuition'' and argumentation about why a metric would be important (and
previous use of such metrics, and - in one isolated case (surprisingingness) -
on a single quote from a mathematician\footnote{Similarly, classification of a
MATHsAiD result as either interesting or uninteresting \cite{roy} was made by
the main system developer: in both cases the main developer had a background in
mathematics.}; and mathematicians themselves, such as Gowers' suggestion that we
can identify features which are commonly associated with good proofs
\cite{gowers}. All of these approaches rest upon the assumption that it makes
sense to speak of ``the good'' in mathematics. This assumption is supported by
concepts such as Erdos's ``The Book'' -- a theoretical construct of a book which
contains the most elegant proof of every theorem (brought to life in
\cite{aigner2010proofs}). However,
empirical studies by psychologists such as Inglis call into question such
assumptions: work by Inglis and colleagues has shown that there is not a single
standard of validity among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
nor beauty. %FIXME needs citation

>>>>

\newcommand{\etal}{{\em et al.}}
\newcommand{\Bagai}{Bagai \etal{}}

@@

A major challenge for both ATF and MTE is choosing how to narrow down
the set of generated conjectures to those deemed ``interesting'', since this is
an imprecise term with many different interpretations. For example, all existing
approaches agree that simple tautologies are ``uninteresting'', but differ when
it comes to more complex statements.  Colton \etal{} give a survey of systems for
theory formation and exploration, and their notions of ``interestingness'' for
concepts and conjectures, identifying six notions in
total~\cite{colton2000notion}. These notions, as applied to conjectures, are:

@@

{\bf Applicability} depends on the number of models in which a
conjecture holds. The system of \Bagai{} conjectures the non-existence of
objects, and hence favours statements with zero applicability. Other systems
treat applicability as a positive aspect: the more applicable the statement, the
more interesting it is.

@@

\begin{table}
  \centering
  \begin{tabular}{ |l|l|c|c|c|c|c|c| }
    \hline
    \multirow{2}{*}{\textbf{Program}}                     &
    \multirow{2}{*}{\textbf{Conjecture Types}}            &
    \multicolumn{6}{c}{\textbf{Interestingness Measures}} \\ \hhline{~~------}
    \tRow{          &                    & \iE & \iN & \iS & \iA & \iC & \iU}
    \tRow{AM        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{GT        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{Graffiti  & \tINE              &   X &   X &   X &     &   X &   X}
    \tRow{\Bagai{}  & \tNE               &     &   X &     &   X &   X &    }
    \tRow{HR        & \tIFF, \tIMP, \tNE &   X &   X &   X &   X &   X &   X}
    \tRow{QuickSpec & \tEQ               &   X &   X &     &   X &   X &    }
    \tRow{Speculate & \tCON \tEQ / \tINE &   X &   X &     &   X &   X &    }
    \tRow{IsaCoSy   & \tEQ               &   X &   X &     &   X &   X &   X}
    \tRow{IsaScheme & \tEQ               &   X &   X &     &   X &   X &   X}
  \end{tabular}
  \caption{Classification of MTE systems from \cite{colton2000notion}, extended
    to those compared in \cite{claessen2013automating} (QuickSpec is the
    conjecture generation component of HipSpec). The interestingness measures
    are \iE{}mpirical plausibility, \iN{}ovelty, \iS{}urprisingness,
    \iA{}pplicability, \iC{}omprehensibility (low complexity) and \iU{}tility.}
  \label{table:colton}
\end{table}

@@

We summarise the criteria in Table 1 and state which were used in key
ATF/MTE systems. The entries for AM, GT, Graffiti, \Bagai{} and HR are based
on this survey (the latter is Colton's own system). We extend this analysis to
some more recent systems in the remaining rows (based on our understanding of
their function): QuickSpec, Speculate~\cite{braquehais2017speculate}, IsaCoSy
and IsaScheme.

@@

%IsaScheme
Monta{\~n}o-Rivas \etal{} have implemented a scheme-based approach to MTE in their
IsaScheme system~\cite{MontanoRivas2011}. Schemes are higher-order formulae
which can be used to generate new concepts and conjectures; variables within the
scheme are instantiated automatically and this drives the invention process.
For instance, in the theory of natural numbers, given the concepts of successor
($suc$), addition ($+$) and zero ($0$), IsaScheme can use the following scheme
to invent the concept of multiplication:

@@

Fundamentally, any mathematical reasoning system must decide on, and formalise,
what counts as ``the good'' in mathematics.  Obvious metrics such as ``true'' or
``provable'' include trivial tautologies, while at the same time failing to
capture the ``almost true'', which can be a valuable trigger for theory change,
as demonstrated by Lakatos in his case studies of mathematical
development~\cite{lakatos}. ``Beautiful'' is another -- albeit vague -- commonly
proposed metric. Methods for finding out what that means have come from
neuro-scientists, such as the study by Zeki \etal{} in which they test whether
mathematicians' experiences of abstract beauty correlates with the same brain
activity as experiences of sensory beauty~\cite{zeki}; computer scientists, such
as Colton's metrics above which are proposed based largely on a computer
scientist's ``intuition'' and argumentation about why a metric would be
important (and previous use of such metrics, and - in one isolated case
(surprisingingness) - on a single quote from a mathematician\footnote{Similarly,
  classification of a MATHsAiD result as either interesting or
  uninteresting~\cite{roy} was made by the main system developer: in both cases
  the main developer had a background in mathematics.}; and mathematicians
themselves, such as Gowers' suggestion that we can identify features which are
commonly associated with good proofs~\cite{gowers}. All of these approaches rest
upon the assumption that it makes sense to speak of ``the good'' in mathematics.
This assumption is supported by concepts such as Erdos's ``The Book'' -- a
theoretical construct of a book which contains the most elegant proof of every
theorem (brought to life in~\cite{aigner2010proofs}). However, empirical studies
by psychologists such as Inglis call into question such assumptions: work by
Inglis and colleagues has shown that there is not a single standard of validity
among contemporary mathematicians %FIXME https://dspace.lboro.ac.uk/dspace-jspui/handle/2134/12150
nor beauty. %FIXME needs citation
